---
title: "MICR 575 Final Project: Structural Equation Models"
output:
  github_document
---


# Introduction:

Below are a series of predictive models known as structural equation models, or SEM(s) for short. These models analyze how observable variables (variables we can directly measure in reality) correlate with variables that cannot be directly measured (latent variables), which, in this instance, are hyporheic zone health which is a combination of geochemical conditions , environmental conditions, microbial populations, and invertebrate populations. The hyporheic zone (HZ) is an ecotone within riparian systems where substantial mixing of surface water and groundwater occurs across fluctuating (hydro)geological, biological, and atmospheric boundaries (Bencala, 2000; Boulton et al., 2010; Ye et al., 2012). The HZ is an important habitat for microbes, lotic macroinvertebrates, and groundwater invertebrates, as well as larger organisms at higher trophic levels (Holmes, 2000). As much as 90%–98% of biogeochemical processing within a watershed occurs within the riparian zone, or the interface between aquatic and terrestrial habitats, which is dominated by the hyporheic zone. The models in this project utilize structural equation modeling within the “lavaan” package to produce path analysis plots which are an examination of the direction and magnitude of the relationship between variables. Structural equation models are used in a variety of modeling situations when typical regression models may not be sufficient. It must be noted, however, that the user must use caution when interpreting SEM results as the model conditions and convergences may not be adequate for causation (Beran & Violato, 2010). These relationships have the potential to reveal hidden processes/relationships between the many variables considered in my research. 

*References:* 

* Bencala K, Walters R (1983). Simulation of Solute Transport in a Mountain Pool-and-Riffle Stream: A Transient Storage Model. Water Resources Research - WATER RESOUR RES 19: 718-724.

* Beran, T.N., Violato, C. Structural equation modeling in medical research: a primer. BMC Res Notes 3, 267 (2010). https://doi.org/10.1186/1756-0500-3-267

* Boulton AJ, Datry T, Kasahara T, Mutz M, Stanford JA (2010). Ecology and management of the hyporheic zone: stream–groundwater interactions of running waters and their floodplains. Journal of the North American Benthological Society 29: 26-40.

* Holmes R (2000). The Importance of Ground Water to Stream Ecosystem Function. pp 137-148.

* Ye S, Covino TP, Sivapalan M, Basu NB, Li H-Y, Wang S-W (2012). Dissolved nutrient retention dynamics in river networks: A modeling investigation of transient flows and scale effects. Water Resources Research 48.

For more information on SEM's and Lavaan follow these links: 

Lavaan package (https://lavaan.ugent.be/) Sem package (https://cran.r-project.org/web/packages/sem/sem.pdf)

```{r}

library(lavaan)  #for performing the analyses
library(semPlot)  #for plotting your analyses
library(dplyr)  #for subsetting data 
library(png) #for displaying .png files
semcsv=read.csv('sem_final.csv')
```

```{r}
head(semcsv)  #view data file to confirm what data is pulled in
```

# Structural Regression Model 
```{r}

#First step is creating the sem model with latent variables
#Second step is creating predictor variables by regression of latent variables

sem.model.measurement <- "CHEM =~ 1*DO + H2S + CH4 + TEMP + COND + NH3N + PH
#make indicator latent environmental health with observable geochemical factors and DO as marker

MIC =~ 1* M.Diversity + M.Abundance
#make indicator latent microbial health factor with biomass and prokaryote diversity as the marker

INV =~ 1* I.Diversity + I.Abundance
#make indicator latent invertibrate abundance (counts) and diversity variable as the marker

ENV =~ 1* DD + O.Cond + G.Cond + K
#make indicator latent environmental health factor with overburden conditions, geological conditions, hydraulic conductivity, with downstream distance variable as the marker

HZHEALTH =~ CHEM + MIC + INV + ENV" 

sem.fit.measurement <- sem(sem.model.measurement, data = semcsv)
summary(sem.fit.measurement, fit.measures = TRUE)

semPaths(sem.fit.measurement, "par", edge.label.cex = .75, edge.label.position= .5, fade = FALSE, layout="tree3", intercepts = TRUE, residuals = TRUE, thresholds = TRUE, intStyle = "multi", combineGroups = TRUE)

#plot the CFA. you can change layout with layout = argument. see ?semPaths() for more. 
```

```{r}
#Key for SEM Path Diagrams

library("png")
pd <- readPNG("PathDiagram.png", native = TRUE)
plot.new() 
rasterImage(pd,0,0,1,1)

```

# Exogenous Model 
```{r}
exmodel <- 'CHEM =~ 1*DO + H2S + CH4 + TEMP + COND + NH3N + PH
#intercepts (nu = tau) 

#Specify parameters on observed variables
DO ~ 1
H2S ~ 1 
CH4 ~ 1
TEMP ~1 
COND ~ 1
NH3N ~ 1
PH ~ 1'

Fitexmodel <- sem(exmodel, data=semcsv) 
summary(Fitexmodel, standardized=TRUE)
semPaths(Fitexmodel, "par", edge.label.cex = 1, edge.label.position= .5, fade = FALSE)
```

# Conclusion
There are a mix of observable variables that seem to have directly positive or negative correlations with my chosen latent variables. However, some of the covariances of my observed variables are extremely high, indicating that the relationship between variables may be very strong, which does not necessarily mean an actual correlation exists. 

Although the analyses were completed I received the following warnings from lavaan: 

lav_data_full(data = data, group = group, cluster = cluster, : lavaan
WARNING: some observed variances are (at least) a factor 1000 times larger than others; use varTable(fit) to investigate
 
Warning in lav_model_vcov(lavmodel = lavmodel, lavsamplestats = lavsamplestats, : lavaan WARNING:
Could not compute standard errors! The information matrix could not be inverted. This may be a symptom that the model is not identified.

This may indicate that data clean up needs to occur, since the variances are very high in some of the observable variables as previouslt mentioned. Moreover, some of the data pulled into these models is not "real" data, instead it is data randomly generated and entered into the model. Data for these models may need to be normalized (z-score) prior to utilizing the SEM(s). 












